---
title: The war on drugs is a war on us - young people who use drugs and the fight for harm reduction in the Global South 
author:
  - name: MJ Stowe
    orcid: 0000-0002-8640-6532
    corresponding: true
    email: mjstowe@kirby.unsw.edu.au
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - South African Network of People Who Use Drugs (SANPUD), South Africa
      - The University of Pretoria, Department of Family Medicine, South Africa
      - The Kirby Institute, UNSW Syndey, Sydney Australia
  - name: Rita Gatonye
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - Women in Response to HIV/AIDS and Drug Addiction, Nigeria
  - name: Ruby Lawlor
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - YouthRISE
  - name: Ishwor Mahajan
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - YouthRISE
  - name: Seyi Kehinde
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - The National Drug and Alcohol Research Centre, UNSW Syndey, Sydney, Australia
  - name: Sidarth Arya
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - State Drug Dependence Treatment Centre, Institute of Mental Health, Pt B.D.S University of Health Sciences, Rohtak, India 
  - name: Jorge Herrera Valderrábano
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - Ágora, Mexico City, Mexico 
  - name: Angela Mcbride
    orcid: 
    corresponding: false
    roles: []
    affiliations:
      - South African Network of People Who Use Drugs (SANPUD), South Africa
  - name: Florian Scheibein
    orcid: 0000-0001-9585-5068
    corresponding: false
    roles: []
    affiliations:
      - School of Health Sciences, South East Technological University, Waterford, Ireland 
  - name: Emmy Kageha Igonya
    orcid: 0000-0001-7155-3786
    corresponding: false
    roles: []
    affiliations:
      - African Population and Health Research Centre, Kenya 
  - name: Danya Fast
    orcid: 0000-0001-5994-3669
    corresponding: false
    roles: []
    affiliations:
      - The British Columbia Centre on Substance Use
keywords:
  - Youth
  - Global South
  - Young people who use drugs
  - Harm Reduction
abstract: |                                                                                          
  In the Global South, young people who use drugs (YPWUD) are exposed to multiple interconnected social and health harms, with many low- and middle-income countries enforcing racist, prohibitionist-based drug policies that generate physical and structural violence and oppression. While harm reduction coverage for YPWUD is suboptimal globally, in low- and middle-income countries youth-focused harm reduction services are particularly lacking. Harm reduction programs in these countries are often powerfully shaped by global health funding regimes that restrict progressive programming and reach. In this commentary we highlight the efforts of young people, activists, allies, and organisations across some Global South settings to enact programs such as those focused on peer-to-peer information sharing and advocacy, overdose monitoring and response, and drug checking. We draw on our experiential knowledge and expertise to identify and discuss key challenges, opportunities, and recommendations for youth harm reduction movements, programs and practices in low- to middle-income countries and beyond, focusing on the need for youth-driven interventions. We conclude this commentary with several calls to action to advance harm reduction for YPWUD within and across Global South settings.

plain-language-summary: |
  While young people who use drugs (YPWUD) are increasingly exposed to related social and health harms, YPWUD in low- and middle-income countries from the Global South often face particular challenges and are at increased risk of mortality and morbidity. The sociopolitical environment characteristic of many LWMICs place many harm reduction programs in situations where they are limited by stringent donor funding, lack of skills transfer and up-scaling and structural violence imposed through the "war on drugs". Despite these challenges, several success stories exist. 
date: last-modified
css: custom01.css
bibliography: references.bib
citation:
  container-title: Young People Who Use Drugs
number-sections: true
always_allow_html: true
---

## Introduction

In the Global South, young people are often exposed to multiple
interconnected social and health harms as a result of the war on drugs
[@stockings2016; @degenhardt2016; @kimmel2020], with many low- and
middle-income countries imposing racist, classist, and prohibitionist
drug policies through a continuum of violence that encompasses physical
and structural assaults [@daniels2021; @tinasti2018; @duarte2020]. Young
people who use drugs (YPWUD) in the context of various intersections of
age, race, class, gender, sexuality, mental health challenges, and
involvement in criminalized income generation activities such as sex
work often experience heightened violence and oppression and worse
health and social outcomes [@jacobs2020; @delatorre-luque2021;
@tinasti2018; @krug2015]. Despite increasing global coverage of harm
reduction services, there remains a lack of youth-focused harm reduction
services, especially in low- and middle-income countries
[@colledge-frisby2023]. In general, across these settings, public health
systems are often characterized by systemic under investment and
deteriorating infrastructures, human resource crises, and corruption
[@serebryakova2021]. Conflicting public health, international donor
funding, and law enforcement agendas impede the implementation of many
evidence-based harm reduction programs. The criminalization and
moralization of drugs and the people who use them powerfully undermines
access to those harm reduction services that do exist [@wodak2012;
@toumbourou2007]. In places where it is possible to access harm
reduction programs (usually run by non-governmental organizations),
YPWUD are disproportionately underserved relative to older populations
in these contexts [@ogundipe2018; @krug2012].

Across the Global South, the war on drugs is too often a violent,
racist, and classist war on YPWUD in the context of multiple and
intersecting forms of oppression and limited access to care
[@daniels2021] . Yet, YPWUD as well as youth-led and youth-inclusive
organizations in these settings are actively pioneering harm reduction
programs to meet their needs and the needs of their communities
[@canêdo2022]. This commentary is authored by YPWUD -- past and present
-- from countries throughout the Global South, alongside academic and
community allies from low- and middle-income countries as well as higher
income countries. We are a heterogeneous group, yet each of us embraces
harm reduction as a set of ideological principles and pragmatic
strategies rooted in social and health justice and a commitment to human
rights, including the right to health for all people [@pauly2008]. We
believe that harm reduction is characterised by an absence of judgement
towards drug use and respect for an individual's choice to use drugs
[@zampini2018]. Since 2009, the World Health Organisation has provided a
list of harm reduction interventions for the prevention, treatment and
care of people who inject drugs and are living with or at risk of HIV,
including needle distribution and exchange programs, opioid agonist
therapy, and HIV testing and treatment programs
[@worldhealthorganization2009]. However, we argue that harm reduction
programming for people who use drugs, including YPWUD, must extend
beyond HIV prevention, testing and treatment. Programming must include
interventions such as the distribution of a range of supplies (not just
needles -- for example, safer smoking kits), take-home naloxone
programs, drug checking services, drug consumption spaces, and peer-led
information sharing, support, and advocacy. Unfortunately, across many
Global South settings, donor funding has been insufficient to support
this kind of comprehensive harm reduction programming, including for
YPWUD [@serebryakova2021]. We have therefore taken matters into our own
hands, in some cases despite tremendous risks to our safety and the
safety of our communities.

The purpose of this commentary is to further ignite much needed
conversations about YPWUD and harm reduction in the Global South. It
centres the diverse efforts of young people, activists, allies, and
organisations across numerous Global South settings to enact programs
such as those focused on peer-to-peer information sharing and advocacy,
overdose monitoring and response, and drug checking. This is by no means
a comprehensive overview of what is happening when it comes to YPWUD and
harm reduction in the Global South. So many success stories are missing
from what follows, in part because we struggled to connect -- and stay
connected with -- young drug user activists and harm reduction
practitioners across the globe. These young people are oftentimes
overworked, overwhelmed, and under compensated as they attempt to keep
themselves and their communities safe while under the weight of poverty,
the drug war, and other forms of oppression. They may be completely new
to these kinds of scholarly outputs, and lack access to mentors who are
able assist with the challenging work of writing (often in a second
language). The timeline of this kind of work can also be frustrating; a
large investment of time and energy is required, but the rewards of
contributing are unclear, especially as time passes and a publication
has still not come to fruition. Sharing a number of the harm reduction
success stories that we were able to collect, we draw on our
experiential knowledge and expertise to identify and discuss key
challenges, opportunities, and recommendations for youth harm movements,
programs and practices in low- to middle-income countries and beyond,
focusing on the need for youth-driven healthcare and harm reduction
interventions.

## A note on language {#sec-language}

There are major limitations to the language of both "young people who
use drugs (YPWUD)" and the "Global South" that we employ throughout this
commentary. Both terms may ultimately obscure more than they reveal,
because they often seem to place finite boundaries around what in
reality are much messier social and geographic categories. Definitions
of "young people" and "youth" -- and lived experiences of these
categories -- vary widely across settings and contexts, where
intersections of age, gender, sexual orientation, class, race, human
immunodeficiency virus (HIV) status, and other dimensions of
positionality are mediated by particular configurations of power and
political economy to shape these social categories and lived experiences
[@duarte2020]. Does it make sense to talk about young people and youth
as those under twenty-four or twenty-nine years of age, when many
individuals continue to strongly identify with these categories -- and
with youth drug user activism and movements -- well into their thirties,
often due to entrenched, shared circumstances of precarity as well as
shared visions for possible solutions [@canêdo2022; @stowe2022]?
Conversely, do age-definitions of young people and youth make sense when
referring to those for whom poverty, lack of education, unemployment,
violence, migration, HIV, and other difficulties have forced them to
move directly from childhood to adulthood, without the possibility of
experiencing youth as a period of transition or activism and power
[@dambruoso2022]? Similarly, it may not make sense to talk about a
Global South composed of countries and regions in Africa, Latin America,
and parts of Asia that meet certain criteria according to the World Bank
income-per-capita index, when many so-called "Global North" settings are
home to populations experiencing similar levels of entrenched poverty
and structural oppression, also as a result of historical and ongoing
processes of colonialism and capitalism [@buxton2020]. Recognizing the
limitations of this language, in this commentary we have chosen to use
the term Global South (rather than listing out a number of discrete
countries and regions in various instances) in order to emphasize some
of the common and disproportionate impacts of the war on drugs on YPWUD
across low- and middle-income countries [@buxton2020]. We use the term
young people to mark ourselves out as a unique demographic with specific
harm reduction priorities, needs, and desires that is simultaneously
characterized by fluidity of meaning and association not necessarily
determined by numerical age.

## Shared Challenges {#sec-challenges}

There is a severe lack of data on the situation of YPWUD in the Global
South [@doyle2022]. What we do know is that those between the ages of 14
and 34 account for more than one third of the population in low-income
countries, and rates of substance use are high and climbing among this
age range[@jumbe2021; @david2023]. YPWUD in the Global South are
vulnerable to a myriad of social and health harms (exacerbated by the
COVID-19 pandemic), including blood-borne infections (e.g. HIV, HCV),
fatal and non-fatal opioid involved overdose, skin and soft tissue
infections, police violence and extrajudicial murder and mass
incarceration [@jumbe2021; @degenhardt2016; @hall2016; @scheim2020].

Global coverage of harm reduction interventions is suboptimal , but this
is particularly the case in lower- and middle-income countries
[@colledge-frisby2023]. For example, the Global State of Harm Reduction
Report [@harmreductioninternational2020] highlights that only fourteen
out of twenty-five countries or regions in these parts of the world have
existing needle exchange programs. Countries or regions that do have
needle exchange programs generally also provide opioid agonist therapy,
such as methadone. However, even in these settings, life-saving
interventions such as drug consumption spaces and take-home naloxone
programs remain largely unavailable [@colledge-frisby2023].

Across Global South settings, YPWUD and the organizations they are a
part of face particularly dire challenges in implementing much needed
harm reduction services due to hostile and militarized governments,
systemic corruption, violent policing, punitive laws and forced
treatment or rehabilitation models, precarious and conditional state and
international funding, and entrenched stigma and oppression
[@dambruoso2022; @desjarlais2013; @krug2012]. High levels of
unemployment, poverty, and homelessness often combine with the
criminalization of drug use (in some cases via the death penalty and
extrajudicial killings), sex work, and sexual and gender identities to
produce egregious human rights violations and make harm reduction
organising and action difficult, and sometimes impossible
[@dambruoso2022; @debeck2017; @rekart2005]. In a system of prohibition,
the Global South is also disproportionately impacted by the various
negative effects of the global demand for illicit drugs, which
particularly impacts countries in Latin America, South East Asia and the
Middle East and North Africa, as well as in transit regions such as West
Africa [@unodc2022].

## Success Stories {#sec-success}

While significant challenges remain for implementing comprehensive harm
reduction programming for YPWUD across Global South settings
[@daniels2021; @chang2021], several of us are actively involved in
implementing various harm reduction programs in our countries. We share
these examples here with the goal of inspiring youth drug user activism
and meaningful policy and programming change across lower- and
middle-income settings globally.

::: panel-tabset
## East Africa

### Empowering young women who use drugs in East Africa through online networking and peer support

A majority of empowerment efforts directed towards women who use drugs
are under-resourced, patriarchal, or fail to consider how complex
intersections of age, gender, class, cultural, and geographic contexts
shape drug use [@lambdin2013]. In 2022, the community-led organization
Women in Response to HIV/AIDS and Drug Addiction (WRADA) set out to
build a network of young women who use drugs in Kenya, Uganda and
Tanzania with support from the International Network of People who use
Drugs. This multi-year community initiative involves bi-monthly online
peer support forums using Zoom. Young women meet on Zoom to discuss and
document regional harm reduction challenges and emerging trends in drug
use and sex work and develop sexual and reproductive health and harm
reduction information tailored to their communities. The project fosters
empowerment by increasing the knowledge, skills and capacities of
participants, growing grassroots harm reduction research and advocacy,
developing context-driven approaches to promoting human rights, and
providing peer mental health support. Despite challenges with internet
connectivity and technological know-how among participants, the project
has resulted in improved relationships between young women who use drugs
and local harm reduction programs across East Africa, greater advocacy
for adoption of best practices through a peer-to-peer learning model,
the generation of more robust evidence for regional harm reduction and
drug policy reform, and increased visibility of young women who use
drugs in East Africa.

## South Africa

### Distributing take-home naloxone kits and overdose education in South Africa

In South Africa, YPWUD are an underserved population frequently exposed
to the social and health harms of HIV and HCV incidence, skin and
soft-tissue infection, poverty, unstable housing and homelessness, and
multiple forms of physical and structural violence [@dambruoso2022;
@peltzer2018]. Heroin use and overdose are increasingly common among
youth [@harker2020]. While South Africa's essential medicines list
includes naloxone for the management of overdose, to date no
state-sponsored naloxone distribution programmes exist [@scheibe2020].
To improve access to this lifesaving overdose antidote among YPWUD and
others, in 2021 the South African Network of People Who Use Drugs
(SANPUD) piloted the country's first (and to date only) take-home
naloxone (THN) program. The program was piloted in Cape Town, Tshwane,
and eThekwini, with two workshops held in each city over a two-week
period. It involved peer-delivered overdose education, including
practical training on the administration of naloxone. Participants
received a naloxone kit with four ampoules of naloxone (0.4mg/ 1ml) and
required medical equipment, as well as a step-by-step guide to
responding to and managing opioid overdoses. The design of these
materials was informed by several community advisory groups that
included youth-led and -focused groups. During the pilot, three opioid
overdoses were successfully reversed. Unfortunately, lack of state
funding and political buy-in has halted the continuation and expansion
of the program. However, the success of the pilot represents an
important moment in the fight against racist and violent drug policies
that continue to criminalize and disproportionately burden YPWUD, and in
particular Black and Brown YPWUD. The successful co-design and delivery
of a youth-led THN program underscores growing calls for non-restrictive
state and NGO funding that can be used to support evidence-based
interventions beyond a narrow set of prescriptive interventions and
programs (e.g., HIV prevention programs focused on people who inject
drugs).

## Nepal

### Training Frontline Workers to Provide Harm Reduction Services to Young People who Inject Drugs in Nepal

In 2019, a government survey revealed that there are over 130,000 people
who use drugs in Nepal, with young people (defined as \< 30 years of
age) accounting for more than two-thirds of this figure [@idpc2021].
Young people who inject drugs in this setting are disproportionately
vulnerable to various social, health, and legal harms [@bhandari2021;
@kakchapati2018]. In response, in 2022 five community organizations (YKP
Lead, Sathi Samuha, Recovering Nepal, Youth RISE, Youth LEAD) came
together to pilot a new training and service delivery project to better
address the needs of young people who inject drugs. As a first step,
focus group discussions with youth and in-depth interviews with service
providers led to the identification of shared problems and possible
healthcare and harm reduction-oriented solutions, including the
identification of key areas for outreach throughout the Kathmandu
Valley. Using this information, twenty-two service providers from
organizations providing harm reduction services in Kathmandu valley were
trained to better understand how social, cultural, political, economic,
geographic and technological contexts affect the practices and service
needs of young people who inject drugs in diverse communities.
Context-informed approaches to care were developed, including online and
peer-to-peer outreach and counselling, expanded hours of operation for
on-the-ground harm reduction service providers, and better referral
mechanisms among providers. The findings and recommendations generated
by the pilot project were later shared with a national audience of
service providers and organizations. The pilot project enhanced
awareness and knowledge among service providers and better equipped them
with skills they need to adopt and deliver harm reduction to young
people who inject drugs.

## Mexico

### Exchanging Knowledge, Building Advocacy, and Challenging Punitive Drug Policies Among YPWUD in Mexico

In Mexico, young people are increasingly exposed to the negative impacts
of the country's militarized and violent approach to addressing drug use
and trafficking, resulting in regular human rights
violations[@hernándezcastillo2019; @morales2020]. Despite these
conditions, international donor funding continues to support "tough on
crime" rhetoric and prioritise law enforcement interventions over
evidenced-based healthcare and harm reduction approaches. In response,
YPWUD have mobilised to hold the now annual Support Don't Punish
Festival as a means of regularly engaging with each other, sharing harm
reduction knowledge and challenging punitive drug policies. The festival
is held each year on June 26th as a community response to commemorate
the UN international day against the illicit trafficking and abuse of
drugs. It is delivered by Instituto RIA and ReverdeSer Colectivo and
provides a safe, non-judgemental space for young people to fight for
their human rights. Festival activities include youth-led marches, harm
reduction information booths, showcases of youth entrepreneurship, and
performances by bands opposed to the oppression of YPWUD. During the
COVID-19 pandemic, the festival transitioned to a virtual event,
expanding its reach to include multiple countries. Throughout the rest
of the year, the festival also organizes other activities around the
country, such as collective murals, art exhibits, social media content,
and harm reduction information materials, including entertaining videos
promoting drug use best practices. The Support Don't Punish Festival has
become a vital platform for harm reduction knowledge exchange, advocacy
and challenging punitive drug policies among YPWUD in Mexico and beyond.

## Columbia

### Drug checking services in Colombia {#sec-drug-checking}

Drug checking services in Colombia have been implemented to decrease
exposure to the growing harms associated with an unregulated drug
supply. In general, drug checking services have multiple aims: to
conduct chemical analysis of substances of submitted directly by the
public; to return results to the service user; to provide a platform for
tailored (rather than general) information exchange between service
users and the service; and to ultimately reduce harms [@barratt2022].
Although reducing drug-related harm via changes in drug using practices
at the point of consumption is key to the success of drug checking,
these services are also highly valued for generating real-time
information about drug market trends that can be actioned rapidly via
text message and social media alerts [@brien2023]. With volunteers and
harm reduction experts located in Bogotá, Medellin, and Cali, the
non-governmental organization Acción Técnica Social implemented a drug
checking service beginning in 2013 through a project entitled Échele
Cabeza Cuando se de en la Cabeza (EC; translated as Use Your Head Before
It Goes to Your Head) [@díazmoreno2022]. Since its inception, EC has
involved YPWUD and used innovative harm reduction communication
strategies to promote self- and community-care: protests, street art,
posters, handouts, flyers, videos, memes, and perhaps most importantly,
maintaining a significant presence on social media. EC began by offering
drug checking services on-site at raves, festivals, and nightlife
events; in 2016, fixed-site drug checking services were introduced
[@díazmoreno2022]. Wherever drug checking services are provided, YPWUD
are provided with tailored information and support backed by scientific
evidence. Based on the testing done across these settings, EC regularly
posts to social media about substances, test results, and alerts,
supporting real time dissemination of critical harm reduction
information. EC has demonstrated the importance of monitoring the drug
market and building online and in-person networks with people who use
drugs, including YPWUD. Unfortunately, despite the effectiveness of this
program and some support from the Mayor's Office of Bogotá and the
Columbian Drug Observatory, financial restrictions continue to limit the
reach of the program [@díazmoreno2022].
:::

## Calls to Action

Drawing on our discussion of shared and challenges and success stories,
we conclude by putting forward seven key recommendations to advance harm
reduction for YPWUD within and across Global South settings:

1.  YPWUD from across the global south must be meaningfully involved in
    harm reduction policy, programming, and activism. We should be at
    the table with government (when this is possible) as well as
    non-governmental organizations and donors when decisions are being
    made. All of the success stories detailed above demonstrate that our
    participation is essential to the development of tailored,
    context-responsive, and effective services and programs that promote
    equity and uphold human rights.
2.  Governments and non-governmental organizations in low- and
    middle-income countries should regularly collect accurate data on
    drug use patterns, including patterns among YPWUD, and use that data
    to inform harm reduction policy, practice, and programming. For
    example, at present, the Government of Nepal only collects data once
    every five to six years, limiting the relevance of this data to
    policy and practice. Yet, the pilot project described above
    demonstrates how up to date information gleaned through focus groups
    and in-depth interviews facilitates the development and adaptation
    of effective harm reduction services.
3.  Financial resources, including international donor and domestic
    funding, must be shifted from punitive law enforcement and drug
    supply reduction towards a continuum of community-based,
    evidence-informed online and on-the-ground harm reduction services,
    including peer-to-peer information generating and sharing programs
    and advocacy networks, take-home naloxone programs, drug checking
    services, and drug consumption spaces.
4.  Government, non-governmental organization, and international donor
    funding focused on harm reduction should be less ties to a narrow
    set of prescriptive interventions and programs (e.g., HIV prevention
    programs focused on people who inject drugs), in order to better
    support the diverse efforts of YPWUD and youth-led and -inclusive
    community organizations to meet their harm reduction needs.
5.  Too often, promising pilot projects end because of a lack of
    sustained funding. Government, non-governmental organizations,
    international donors, and civil society should work to identify and
    scale up promising pilot projects undertaken by and with YPWUD. The
    focus of those providing funding should be on what is happening and
    working on the ground -- and online -- among YPWUD. Recognize that
    funding these projects may produce better results than campaigns and
    programs that are overly general and imposed from the top down.
6.  Towards this end, there must be better coordination and
    collaboration between government, non-governmental organizations,
    international donors, and civil society.
7.  There must be greater efforts to build capacity among YPWUD in the
    global south to undertake harm reduction-focused research, including
    the evaluation of their own programs. Many YPWUD and their mentors
    have tremendous experiential knowledge regarding drug use and harm
    reduction programs, practices, and needs in their communities, but
    lack the ability to translate that knowledge into traditional
    scholarly outputs, including peer-reviewed publications and grants.
    Those conducting funded research -- and building their careers -- in
    these settings who do have these skills, namely, many academics
    based in Northern universities, must commit to doing some of this
    capacity building work together with YPWUD. This can take the form
    of an investment of significant time: co-authoring publications and
    grants (as author Danya Fast did with this piece). It can also take
    the form of a financial investment: YPWUD should be adequately
    compensated for their time and expertise when working on these kinds
    of projects.

## Acknowledgements

The authors acknowledge Alissa Greer (Simon Fraser University, British
Columbia, Canada) and Gloria Lai for their input on the manuscript.

## Declarations

The authors declare that they have no known competing financial
interests or personal relationships that could have appeared to
influence the work reported in this paper.

## Role of the funding source

The open access fee for this paper was covered through a Youth RISE
project which was part of the 4Youth Consortium and funded by the Robert
Carr Fund. DF is supported by a Micheal Smith Health Research BC Scholar
Award. The funders had no role in developing the ideas and information,
structuring the article and identifying the examples within the article.
MJS, FS and DF were responsible for the decision to submit this article
for publication.

## Supplementary Material {#sec-supp-material}

```{r}
# Load the library
library(leaflet)
library(mapview)
# for loading our data
library(sf)
# for plotting
library(lattice)
library(leafpop)
library(mapview)
library(RColorBrewer)
library(viridis)

# Download the shapefile. (note that I store it in a folder called DATA. You have to change that if needed.)
# download.file("http://thematicmapping.org/downloads/TM_WORLD_BORDERS_SIMPL-0.3.zip" , destfile="~/GitHub/YPWUD/world_shape_file.zip")
# You now have it in your current working directory, have a look!

# Unzip this file. You can do it with R (as below), or clicking on the object you downloaded.
# system("unzip ~/GitHub/YPWUD/world_shape_file.zip")
#  -- > You now have 4 files. One of these files is a .shp file! (TM_WORLD_BORDERS_SIMPL-0.3.shp)

shape <- read_sf(dsn = "~/GitHub/YPWUD/world_shape_file/", layer = "TM_WORLD_BORDERS_SIMPL-0.3")

ypwud <- read.csv("~/GitHub/YPWUD/ypwud_locations.csv")



ypwud <- read.csv("~/GitHub/YPWUD/ypwud_locations.csv") %>%
  st_as_sf(coords = c("Longitude", "Latitude"), remove = FALSE)

my.pal2 = brewer.pal(n=7, "GnBu")



# save the widget in a html file if needed.
# library(htmlwidgets)
# saveWidget(m, file=paste0( getwd(), "/HtmlWidget/backgroundMapTile.html", width="1000px"))
```

```{r fgb=FALSE, warning=FALSE, echo=FALSE, include = TRUE}

# Interactive map of ETHOS locations across Australia 

  mapview(shape, alpha.regions = 0, colour = "red", lwd = 2, layer.name = "Country") + 
  mapview(ypwud, zcol="harm_reduction", layer.name = "Harm Reduction Interventions", col.regions = my.pal2) 
 # mapview(ethos.loc, zcol="recruitment", layer.name = "No. of participants recruited")

```

\pagebreak

## References

\singlespacing

::: {#refs}
:::

\pagebreak


```{r}


# Load the qrcode package
# library(qrcode)
# library(officer)
# 
# 
# 
# qr_code("https://mx-jx.github.io/YPWUD/")
# qr <- qr_code("https://mx-jx.github.io/YPWUD/", ecl = "M")
# qr
# plot(qr,  col = c("black", "orange"))

# # Generate the QR code
# qrcode_data <- qr_code(pdf_url)
# 
# # Save the QR code as an image file
# png("ypwud_qr.png")
# 
# 
# # Generate the Q ciss
# plot(qrcode_data, main = "", fg = "white", bg = "black", quiet = TRUE)
```

```{r, echo =  FALSE}
# 
# plot(qr,  col = c("black", "orange"))
```







